---
title: "California School Performance"
output: html_notebook
---

What school characterists are most important towards preparing our children for the college?

```{r message = FALSE}
library(tidyverse)
library(readxl)
```

#Download and label datasets
```{r}
urls <- "http://www3.cde.ca.gov/researchfiles/satactap/sat17.xls"
urls <- c(urls, "https://www.cde.ca.gov/SchoolDirectory/report?rid=dl1&tp=xlsx&ict=Y")
urls <- c(urls, "http://dq.cde.ca.gov/dataquest/dlfile/dlfile.aspx?cLevel=School&cYear=2016-17&cCat=FEP&cPage=filesfepsch")
urls <- c(urls, "http://dq.cde.ca.gov/dataquest/dlfile/dlfile.aspx?cLevel=School&cYear=2016-17&cCat=Enrollment&cPage=filesenr.asp")
urls <- c(urls, "https://www.cde.ca.gov/ds/sd/sd/documents/frpm1617.xls")

files <- c("sat.xls", "schools.xlsx", "languages.txt", "demographics.txt", "poverty.xls")

for (i in 1:length(urls)){
  if (!file.exists(files[i])){
    download.file(urls[i], destfile = files[i], method = "libcurl")
  }
}
```

#Import data sets
```{r message = FALSE, results='hold'}
sat <- read_excel("sat.xls")
schools <- read_excel("schools.xlsx", skip = 5)
languages <- read_tsv("languages.txt")
demographics <- read_tsv("demographics.txt")
poverty <- read_excel("poverty.xls",sheet = 2, skip = 1)
```

#Inspect and clean datasets
We'll start with the `SAT` dataset, as this contains our metrics of performance and is of primary interest in this study
```{r}
database <- list(sat, schools, languages, demographics, poverty)
names(database) <- map(str_split(files, "\\."),1)

str(database$sat)
```

Quick inspection shows that there are redundant entries for districts, counties, and the entire state in addition to entries for individual schools. Since we are interested in features of individual schools, we will remove these other entries. Schools are identifed by their CDS (County-District-School) code, which we will ultimately join these datasets on this code. 
```{r results = 'asis'}
schools <- schools %>% 
  rename(CDS = CDSCode)

sat <- sat %>% 
  filter(rtype == "S") %>% 
  rename(CDS = cds)

sat %>%   
  inner_join(schools, by = "CDS") %>% 
  select(CDS, contains("Benchmark")) %>% 
  head(25)
```

Further inspection shows that there are two types of missing values for regarding school SAT performance: schools in which no students took the SAT (indicated by a value of 0 for `NumTstTakr` and "N/A" in columns related to test performance) and those schools which do not report their performance (indicated with an asterisk, "*"). Are there any common traits to these schools, which might explain these missing values or indicate how to handle these missing values?
```{r results = 'show'}
sat %>% 
  inner_join(schools, by = "CDS") %>% 
  mutate(missingValues = (NumCurrElaBenchmark  %in% c("N/A","*"))) %>%
  select(missingValues,Charter) %>% 
  table()
```
It appears that an unusually high percentage (`r 236/(317+236)*100`%) of those schools with missing values are charter schools. Put another way, charter schools are nearly twice as likely to report missing values as non-charter schools (`r 236/(236+223)*100`% versus `r 317/(1048+317)*100`%). This may make comparisons of relative performance between charter and non-charter schools a tenuous proposition. Since there is no obviously reasonable method for imputing these values, for the momoment, we will simply ommit them for the sake of simplicity. We also convert our "Benchmark" and other enrollment features to numerical values for later computation.

```{r}
sat <- sat %>% 
  filter(!(NumCurrElaBenchmark %in% c("N/A", "*"))) %>% 
  mutate_at(vars(enroll12, NumTstTakr, contains("Benchmark")),funs(as.numeric))
```

We now examine the `schools` dataset for missing values and data-type issues:
```{r}
schools %>% 
  na.omit() %>% 
  anti_join(schools)

head(schools)
```
Initial inspection shows there are (formally) no erroneous missing values in the dataset, but there are entries explicitly marked as "No Data", which we can replace with `NA`'s by re-importing the dataset by flagging these entries.

```{r}
schools <- read_excel("schools.xlsx", skip = 5, na = "No Data")
```

```{r}
sum(is.na(schools$CDSCode))
```

There are nearly 18,000 schools tabulated in this dataset, significantly more than the number of schools reporting SAT performance. Inspection of the dataset's data dictionary shows that the dataset includes information on ALL schools in California, including preschool, elementary, junior high, and adult education schools in addition to high schools. The dataset even includes schools which are no longer operational. All of these additional schools will be neglected when we eventually join it with the `sat` dataset, based on their CDS code, for which there are (thankfully) no missing values in either dataset.
Other possible school features of interest include:
*City: which cities have the highest performing schools? Do schools in larger cities perform better on average than in smaller cities? Can we estimate city size by using the number of schools in that city as a proxy? Alternatively, we could look for another dataset with actual population counts.
*Open/CloseDate: Does students at younger (and presumably more modern) schools than at older schools? Ostensibly, younger schools should have relatively newer equipment and tools and may be more likely to employ experimental or progressive teaching methods. How old is the average school in California? What is the average lifespan of a school? 
*Virtual: The advent of the internet allows for remote education. Do schools that utilize such technologizes just as effective as traditional schools?
*Magnet: Magnet schools offer specialized curricula and draw their students from across their entire school district or even other districts. Do these schools perform better than non-magnet schools? When did magnet schools start to proliferate in California?
*Latitude/Longitude: Where are the highest performing schools? Do schools in close proximity to each other perform similiarly on the SAT? We also note that these geospatial coordinates were improperly imported as character strings instead of numbers and must be converted.

```{r}
schools <- mutate_at(schools, vars(contains("tude")), funs(as.numeric))
```
There are quite a few missing values in the dataset, but it is likely many of them correspond to schools we aren't interested in (junior high, elmentary, closed, etc.). We will put off cleaning these missing values until we've merged with the `sat` dataset and we can determine how much they may affect our analysis.

